### 2012-10-10 4.tex
Writing on front matter, chapters 1-4

Early work in generative phonology recognized that neutralizing, surface-true phonological alternations impose constraints on surface forms and underlying representations (\citealp[283]{Anderson1974}, \citealp[382]{SPE}, \citealp[205f.]{Dell1973}, \citealp[22f.]{SPR}, \citealp{Kenstowicz1977}: chapter 3, \citealp[28f.]{Stampe1979}, \citealp[410f.]{Stanley1967}). Despite isolated claims to the contrary (e.g., \citealp[297]{Hale1965}; \citealp[212f.]{Postal1968}), however, the \emph{communis opinio} (including many of the authors cited above) holds that not all phonotactics---that is, language-specific sound sequence co-occurrence restrictions---can be derived in this fashion, and phonotactic knowledge is independent of the phonology.

@@

@@

\begin{example}[\textsc{Obstruent Voice Assimilation}]
\xymatrix@R=24pt@C=24pt{
\txt{[α \textsc{Voice}]}\ar@{-}[d]\ar@{--}[dr] &                               \\
\txt{C}\ar@{-}[d]                              & \txt{C}\ar@{-}[d]             \\
\txt{[$-$\textsc{Sonorant}]}                  & \txt{[$-$\textsc{Sonorant}]} \\
}
\end{example}

@@

There are two possible analyses of the small number of roots that violate the derived generalization. Some hetero-voiced obstruent clusters, like the one in \emph{ja}[k.d]\emph{aw}, for instance, might be analyzed as compounds by native speakers, and this analysis might itself be driven by the rarity of morph-internal adjacent hetero-voiced obstruents (cf. \citealp[546]{Rice2009d} on compounds in Slave). In fact, \citet{Mattys2001b} find that 9-month-old infants are sensitive to the contrast between [f.t] and exceptional (and unattested) [v.t], treating the latter as an indication of a word boundary and this heuristic is also available to adults \citep{Brown1956,McQueen1998}. Less explanatorily, exceptional roots can be marked [$-$\textsc{Obstruent Voice Assimilation}].

@@

\begin{example}[\textsc{Nasal Place Assimilation}]
\xymatrix@R=24pt@C=24pt{
                          & \txt{\textsc{Place}}\ar@{-}[d]\ar@{--}[dl] \\
\txt{C}\ar@{-}[d]         & \txt{C}\ar@{-}[d]                          \\
\txt{[$+$\textsc{Nasal}]} & \txt{[$-$\textsc{Sonorant}]}              \\
}
\end{example}

@@

@@

Sparse distributions characterize the frequencies of many linguistic objects, from syntactic rules \citep{Yang2009} to word \citep{Teahan1998,Baroni2009} and phoneme \citep{Belevitch1956,Daland2011a} $n$-grams, but are also found in non-linguistic symbol systems \citep{Mandelbrot1954,Miller1957,Chomsky1958,Sproat2010} and randomly generated texts \citep{Miller1957,Li1992}. The importance of sparsity in this context is that it entails a long right tail, making it difficult to determine on statistical grounds alone which unobserved events are impossible and which are accidental gaps.

>>>>

Early work in generative phonology recognized that neutralizing, surface-true phonological alternations impose constraints on surface forms and underlying representations (\citealp[283]{Anderson1974}, \citealp[382]{SPE}, \citealp[205f.]{Dell1973}, \citealp[22f.]{SPR}, \citealp{Kenstowicz1977}: chapter 3, \citealp[28f.]{Stampe1973}, \citealp[410f.]{Stanley1967}). Despite isolated claims to the contrary (e.g., \citealp[297]{Hale1965}; \citealp[212f.]{Postal1968}), however, the \emph{communis opinio} (including many of the authors cited above) holds that not all phonotactics---that is, language-specific sound sequence co-occurrence restrictions---can be derived in this fashion, and phonotactic knowledge is independent of the phonology.

@@

%\citet[208]{Wetzels2001}, cite English as an example of a language without ``general devoicing or assimilatory effects''
%\citet[?]{Lombardi1999} says that English lacks devoicing in Level II.

@@

\begin{example}[\textsc{Obstruent Voice Assimilation}]
%\xymatrix@R=24pt@C=24pt{
%\txt{[α \textsc{Voice}]}\ar@{-}[d]\ar@{--}[dr] &                               \\
%\txt{C}\ar@{-}[d]                              & \txt{C}\ar@{-}[d]             \\
%\txt{[$-$\textsc{Sonorant}]}                  & \txt{[$-$\textsc{Sonorant}]} \\
%}
$\begin{bmatrix} -\textsc{Son} \end{bmatrix}~\goesto~\begin{bmatrix} =\textsc{Voi} \end{bmatrix}~/~\gap~\begin{bmatrix} =\textsc{Voi} \\ -\textsc{Son} \end{bmatrix}$
\end{example}

@@

There are two possible analyses of the small number of roots that violate the derived generalization. Some hetero-voiced obstruent clusters, like the one in \emph{ja}[k.d]\emph{aw}, for instance, might be analyzed as compounds by native speakers, and this analysis might itself be driven by the rarity of morph-internal adjacent hetero-voiced obstruents (cf. \citealp[546]{Rice2009d} on compounds in Slave). In fact, \citet{Mattys2001b} find that 9-month-old infants are sensitive to the contrast between [f.t] and exceptional (and unattested) [v.t], treating the latter as an indication of a word boundary and this heuristic is also available to adults \citep{Brown1956,McQueen1998b}. Less explanatorily, exceptional roots can be marked [$-$\textsc{Obstruent Voice Assimilation}].

@@

% mention Jensen here

\begin{example}[\textsc{Nasal Place Assimilation}]
%\xymatrix@R=24pt@C=24pt{
%                          & \txt{\textsc{Place}}\ar@{-}[d]\ar@{--}[dl] \\
%\txt{C}\ar@{-}[d]         & \txt{C}\ar@{-}[d]                          \\
%\txt{[$+$\textsc{Nasal}]} & \txt{[$-$\textsc{Sonorant}]}              \\
%}
$\begin{bmatrix} $+$\textsc{Nas} \end{bmatrix}~\goesto~\begin{bmatrix} =\textsc{Lab} \\ =\textsc{Cor} \\ =\textsc{Dor} \end{bmatrix}~/~\gap{}~\begin{bmatrix} =\textsc{Lab} \\ =\textsc{Cor} \\ =\textsc{Dor} \\ -\textsc{Son} \end{bmatrix}$
\end{example}

@@

\begin{example}[\textsc{Degemination}]
$\begin{bmatrix} =\textsc{Lab} \\ =\textsc{Cor} \\ =\textsc{Dor} \\ =\textsc{Son} \\ =\textsc{Nas} \\ =\textsc{Cont} \end{bmatrix}~\goesto~\zero~/~\gap~\begin{bmatrix} =\textsc{Lab} \\ =\textsc{Cor} \\ =\textsc{Dor} \\ =\textsc{Son} \\ =\textsc{Nas} \\ =\textsc{Cont} \end{bmatrix}$
\end{example}

% consider what coda-final consonants there are

@@

Sparse distributions characterize the frequencies of many linguistic objects, from syntactic rules \citep{Yang2011b} to word \citep{Baroni2009} and phoneme \citep{Belevitch1956,Daland2011a} $n$-grams, but are also found in non-linguistic symbol systems \citep{Mandelbrot1954,Miller1957,Chomsky1958,Sproat2010} and randomly generated texts \citep{Miller1957,Li1992}. The importance of sparsity in this context is that it entails a long right tail, making it difficult to determine on statistical grounds alone which unobserved events are impossible and which are accidental gaps.

### 2012-10-10 abs.tex
more writing
Writing on front matter, chapters 1-4

This dissertation considers two types of ``gaps'' in the lexicon. In both cases, it is argued that current generative models can be modified minimally to account for the apparent overgeneration of gapped structures. The first part of the dissertation considers the underlying phonological structure of morphs, specifically medial consonant clusters in English. A number of prior studies have argued that the grammar must have a separate ``phonotactic'' module to account

at the level of prosodic word. In both cases, it is argued that the i

>>>>

This dissertation argues for that phonotactic knowledge is primarily derived not from static generalizations but rather from phonological alternations and prosodic parsing. Chapter \ref{intro} outlines the proposal, historical precursors, and predictions. 
%especially with regards to order of acquisition.

Chapter \ref{gradience} considers evidence from wordlikeness rating tasks. It is argued that intermediate well-formedness ratings are obtained whether or not the categories in question are graded. A primitive categorical model of wordlikeness using prosodic representations is outlined, and shown to predict English speakers' wordlikeness judgements at least as acccurately as state-of-the-art gradient wellformedness models. Once categorical effects are controlled for, gradient models are uncorrelated with well-formedness ratings.

Chapter \ref{turkish} considers the relationship between lexical generalizations, phonological alternations, and speakers' nonce word judgements with a focus on Turkish vowel patterns. It is shown that even exception-filled phonological generalizations have a robust effect on wellformedness judgements, but that statistically reliable phonotactic generalizations go unlearned when they are not derived from phonological alternations.

Chapter \ref{clusters} investigates the role of phonological alternations in constraining lexical entries, focusing specifically on medial consonant clusters in English. Static phonotactic constraints previously proposed to describe gaps in the inventory of medial clusters are shown to be statistically unsound, whereas phonological alternations impose robust restrictions on the cluster inventory. The remaining gaps in the cluster inventory are attributed to the sparse nature of the lexicon, not static phonotactic restrictions.

### 2012-10-10 ack.tex
Writing on chapters 2, 4
Writing on front matter, chapters 1-4

So many people contributed to this dissertation that to name them all would compromise the honor of the most worthy. 
More than anyone else, I owe a debt of gratitude to my advisor, Charles Yang, whose influence can be seen throughout. 
Gene Buckley, Mark Liberman, and Rolf Noyer, my commitee members, are responsible for creating an environment at Penn where a student could write this admittedly contenious dissertation. 
And though his name appears nowhere else on this document, it is hard to imagine this work could exist without Bill Labov.

My journey in linguistics begins with the mentoring of Jennifer Cole. 
Steve Anderson graciously agreed to serve on my committee. 
Morris Halle told me to never apologize.

I would like to thank my fellow students at Penn, but especially my cohort, Toni Cook, Laurel MacKenzie, and Stefanie Brody. 
Daniel Ezra Johnson fearlessly proofread (\emph{proofreaded}?) the entire document. 
Maryam Bakht kept me centered.

My gratitude also goes out to the administrative assistants at 3401 Walnut and 255 S. 36th, Lauris Olson at the Van Pelt Library, Katherine McEldoon for a helpful review of numerical cognition research, and Carly Fisher for her unflagging support and friendship.

>>>>

So many people contributed to this dissertation that to name them all would compromise the honor of the most worthy. More than anyone else, I owe a debt of gratitude to Charles Yang, who has done (and put up with) so much for me. I am grateful to my committee members, Gene Buckley, Mark Liberman, and Rolf Noyer, for creating a nurturing environment for quantitative and theoretical approaches to meet. And though his name does not appear again in this document, so much here has been inspired by Bill Labov.

The faculty, staff, and students of the Institute For Research in Cognitive Science have provided me with not merely an office, but a home, complete with dated early-90s decor. Thanks to my officemates, Chandan Narayan, John Trueswell, Laurel Sweeney and the rest of the staff there, as well as Amy Forsyth back at Williams Hall.

So many of my fellow students at Penn have had a hand in this document, none so much as Toni Cook and Laurel MaKenzie, who have supported me so much over the past few years. Without Joe Fruehwald, Marielle Lerner, Constantine Lignos, Brittany McLaughlin, and Hilary Prichard, it wouldn't have been nearly as much fun. Carly Fisher and Katherine McEldoon have provided me with unflagging support from afar.

The ideas I first encountered under the tutelage of Jennifer Cole and Richard Sproat at the University of Illinois continue to guide my research. Insofar as I have acknowledged historical precursors, it is thanks to Steve Anderson, who graciously agreed to serve on my committee. Morris Halle taught me to speak clearly and to never apologize.

### 2012-10-10 pre.tex
Writing on chapters 2, 4
Writing on front matter, chapters 1-4

Perhaps the most important fact for the study of speakers' internal knowledge of language is Wilhelm von Humboldt's observation that speakers make ``infinite use of finite means'' CITE \citep{Chomsky1966}.

An enduring demonstration of ``infinite use'' is the famous emph{wug-test} \citep{Berko1958}. Jean Berko-Gleason introduced English children aged ?-? to pictures of concrete objects (such as a small bird), labeled them with a nonce word (``that's a \emph{wug}''), and then asked them to finish the sentence

The children in the study produced the same response you may have: the plural is \emph{wugs} (where $\langle$s$\rangle$ writes a final [z]).

Of course, children have never heard the plural of these nonce words, but they extend /-z/ to these new words.

But what of the ``finite means'' from which grammars spring? On this, a good deal less has been said.
a study of spaekers' knowledge of natural
The most important fact for

In this dissertation, I critique two such epicyclic proposals, and present solutions using independently-motivated grammatical machinery to account for the data.

The first half of this dissertation began with a suggestion from Charles Yang that the sparse nature of the lexicon might itself account for at least some phonotactic gaps. The success of the resulting study, in chapter four, led me to question the received wisdom that speakers' knowledge of phonotactic generalizations is gradient and potentially independent of the phonological system.

These results support a null hypothesis in which phonotactic knowledge is derivative of the language acquisition device and phonological processes. While I present this theory as a falsifiable scientific hypothesis, an alternative would treat this proposal as a \emph{negative heuristic}, a

guides us away from ``phonotactic grammar'' as an explanatory device.

even those who are not inclined to accept this proposal should note how much of the coverage traditionally attributed to

>>>>

In a justly-famous study, \citet{Pierrehumbert1994} notes the general absense of velar-labial consonant clusters in English (though note the existence of a few exceptions like \emph{pigment} and \emph{rugby}). When I first encountered this study, as an undergraduate, I simply could not integrate this arbitrary gap with a view of the language faculty as the inexhaustable fount of productivity responsible for such wonders as [wʌgz] and \emph{Colorless green ideas sleep furiously}. This shocking observation was filed away for further meditation, in the hope that one day I might grok.

The atoms of linguistic theory tend to be sparsely distributed, with ``few giants'' (/t/, \emph{the}) and ``many dwarves'' (/ʒ/, \emph{zymurgy}) in any sample. Sparsity is exacerbated for complex objects: however common \emph{the} is, any two word sequence containing it will be far less frequent. This dissertation began in earnest with a suggestion from Charles Yang that perhaps sparsity---the missing and rare data which have long plagued statistical approaches to language learning and processing---might be the only reason, e.g., English has no words containing a [k.p] cluster.

Linguists-in-training are inculcated, early and often, with an respect for productivity---the ``infinite use of finite means'' \citep[8]{ASPECTS}---over all else. Psycholinguists have been more free to pursue less combinatorial, less granular approaches to language. Many developmentalists have argued that the impoverished nature of young children's speech reflects whole phrases stored in memory, not phrase structure rules. It is therefore informative that the relatively sparse distribution of phrases in child speech is a practical \emph{necessity} of generativity and sparsity \citep{Yang2011b}. Natural language engineers, unburdened by the need to construct models which conform to the cognitive constraints imposed by human minds, have endowed statistical parsers with phrasal storage mechanisms ``bilexical'' rules \citep[e.g.,][]{Collins1999}, but these offer little no additional coverage \citep{Bikel2004}. I suspect this is not because stored phrases are inherently uninformative, but rather because only a tin small number of two-word sequences can be reasonably expected to occur with sufficient frequency in any corpus.

Similarly, theorists of lexical processing postulate that complex words are stored in memory to account for the facilitatory effect of whole word frequency on reaction times to complex words \citep[e.g.,][]{Alegre1999,Baayen1997a,Sereno1997}. But this effect of word frequency obtains regardless of whether or not whole word storage is postulated \citep{Taft2004b,LignosInPress}. Furthermore, in languages with rich agglutinative morphology, the role played by a whole word storage mechanism, should one be evidenced, would necessarily be quite marginal \citep{Hankamer1992,Chan2008}. According to one estimate, children acquiring American English hear 165,000 word tokens a week \citep{Swingley2007}. At that rate, it would take more than two months for a child to encounter the 1.5 milllion inflectional variants of a single verb in Archi \citep[467]{Kibrik1998},
%, cited in \citealt[122]{Bauer2001}), 
even if adult speech consisted of nothing but orderly recitation of the inflectional forms of this one verb, one after the other, without repetition.

By
some

I can only hope that this humble theory has even a fraction of the influence on phonotactic thinking that

There are two ways to read this thesis. First, it can be regarded as evidence for a particular, falsifiable theory of phonotactic learning and knowledge. If, for instance, some future study finds that English speakers rate nonce words containing the unattested [k.p] cluster to be less well-formed than similar words containing [k.t], a cluster found in dozens of English words, then this theory will need to be seriously revised. Suggestive evidence comes an auditory wordlikeness task performed by \citet[782]{Munson2001} which includes this contrast. Speakers rated [bɪk.pɘt] and [fæk.pɘt] slightly more English-like than [bɪk.tɘp] and [fæk.tɘp], respectively, and drew no contrast between the pair [nɪk.pɘt]-[nɪk.tɘp].

Other claims are not essential to the theory here, but are rather principled null hypotheses, and evidence to the contrary would result only small modifications to the theory. For instance, it is not claimed that wordlikeness judgements are inherently binary, but rather that the evidence currently available is insufficient to justify the considerable additional complexities of gradient wellformedness models.

Under another reading, the claims here represent a \emph{negative heuristic} guiding the phonologist away from static phonotactics as an explanatory device. Then this work is a claim about something grammars do not contain, not a theory of what they do, much like \emph{Richness of the Base} (RotB), an important intellectual precursor to this work (see \citealt[6]{Bye2001} and \citealt[73]{PE} for further discussion). \emph{Lexicon optimization}, however, is an example of a theory which produces grammars which conform to RotB.

A strict application of this negative heuristic produces some interesting predictions. If speakers are shown to have internalize a phonotactic generalization, then there must be a phonological process to blame. One clear case where processes, and not merely restrictions on underlying representations, are implicated is in Quecha, where ejective co-occurrence restrictions are reflected in speakers' difficulties in producing nonce words containing multiple ejectives \citep{GallagherInPress}.

This negative heuristic also allows for the possibility of using psycholinguistic evidence to decide between competing phonological analyses. For instance, there is a long-standing debate concerning the representation of Sanskrit aspiration alternations like those represented by \emph{bodhati}-\emph{bhotsyati} `he wakes-he will wake'. One analysis, going back to the ancient grammarian Pāṇini \citep{Sag1974} and to \citet[\S141f.]{Whitney1889}, posits an underlying representation /budh-/ with aspiration shifting leftward in certain contexts: modern proponents of this analysis include \citet{Borowsky1983}, \citet[][59f.]{Hoenigswald1965}, \citet{Kaye1985}, \citet{Sag1974,Sag1976} \citet{Schindler1976}, \citet{Stemberger1980}, and \citet[][\S141f.]{Whitney1889}. An alternative analysis posits an underlying representation /bhudh-/ and an apsirate dissimilation process (\citealt{Anderson1970}, \citealt{Hoard1975}, \citealt[\S3.2]{Kiparsky1965}, \citealt{Phelps1973}, \citealt{Phelps1975b}, \citealt[109f.]{Zwicky1965}). It is not my intention to adjudicate between these competing analyses, or to evaluate the claim that both accounts are deeply flawed \citep{Janda1989}, but simply to note that they make different predictions under the theory proposed here. While the former, aspirate shift analysis denies the existence of diaspirate roots, this fact lacks synchronic motivation, so the co-occurrence restriction may go unlearned. On the other hand, under the aspirate dissimilation account, underlying diaspirate roots exist but speakers internalize synchronic processes responsible for their absence on the surface.

%\citet[38f.]{MacEachern1999} discusses a similar restriction on co-occurign aspirate roots in Ofo, but also notes, following \citet{DeReuse1981}, that this generalization holds both within roots and also within complex words. For instance, the aspiration of ['osk\asp{}a] `crane' disappears when combined with [a'f\asp{}ã] `white': [oskaf\asp{}ã] `white or American egret'. Under the negative heuristic, the root co-occurrence restriction and this alternation are both the result of a phonologically general process of aspirate dissimilation.
